---
title: "Code Through - Whole Game (Hadley Wickham)"
# output: hrbrthemes::ipsum_pdf
output: github_document
---



# US building permits

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE,  
                      tidy = FALSE,
                      size = "small",
                      dev = "cairo_pdf")
library(tidyverse)
library(magrittr)
library(hrbrthemes)
library(skimr)
library(Cairo)
library(extrafont)
extrafont::loadfonts()
```

# Import

Download the data from [this site](https://www.recenter.tamu.edu/data/building-permits#!/state/United_States)

```{r import}
# fs::dir_ls("data")
permits_raws <- read_csv("data/MSAdataPermit_fullmo.csv")
permits_raws
```

* area = metropolitan standard area 
* date = month / year (character vector, or strings)  
* f1 = 1 family, f24 = 2-4 families, f5 = 5+ families  
* units = number of buildings, change in units; value = average value of building; valchange = change in value 

```{r permits}
permits_raws %>% 
  tidyr::separate(data = ., 
                  col = date, 
                  into = c("month", "year"), 
                  sep = "/", 
                  convert = TRUE) %>% head()
permits <- permits_raws %>% 
  tidyr::separate(data = ., 
                  col = date, 
                  into = c("month", "year"), 
                  sep = "/", 
                  convert = TRUE)
```

## Basic EDA

These are just counts of the categorical data

```{r EDA}
permits %>% dplyr::count(year) %>% utils::head()
permits %>% dplyr::count(area) %>% utils::head()
permits %>% dplyr::count(area) %>% dplyr::count(n)
```

## Create `date` variable

This will put time on the x-axis. January 2007 will be `2007`, February 2007 with be `2007` plus a little bit more. 

```{r create_date}
permits <- permits %>% 
  dplyr::mutate(date = year + (month - 1) / 12)
permits %>% dplyr::glimpse(78)
```

First plot with points. 

```{r point}
permits %>% 
  ggplot2::ggplot(aes(x = date, y = f1units)) + 
    ggplot2::geom_point()
```

Not very useful. Try lines.

```{r line}
permits %>% 
  ggplot2::ggplot(aes(x = date, y = f1units)) + 
    ggplot2::geom_line()
```

Not useful--needs the `group = area`.

```{r line_grouped}
permits %>% 
  ggplot2::ggplot(aes(x = date, y = f1units)) + 
    ggplot2::geom_line(aes(group = area))
```

> I'm going to limit my plot to 2005 - 2017 to look more like the plot in the video. 

## Filter data to 2005 - 2017

```{r filter}
permits <- permits %>% dplyr::filter(date >= 2005 & date <= 2017)
```

Now do this plot again. 

```{r line_grouped_v2}
permits %>% 
  ggplot2::ggplot(aes(x = date, y = f1units)) + 
    ggplot2::geom_line(aes(group = area))
```

## Focus 

Focus on big cities. This might bias results (big cities are different than small cities).

```{r f1units}
f1units <- permits %>% 
  dplyr::group_by(area) %>% 
  dplyr::summarise(mean = mean(f1units)) %>% 
  dplyr::arrange(desc(mean))
f1units %>% head(10)
```

Now filter this to those greater than 100 every month. 

```{r 100_mean}
f1units %>% 
  # how many cities have a mean f1units over 100?
  dplyr::filter(mean > 100) %>% 
  # use this to look at the structure of resulting 116
  dplyr::glimpse(78)
```

## Semi join to `permits`

> this is a cool trick to get the cities with a mean > 100 (in `f1units` data frame).

```{r semi_join}
permits_big <- permits %>% 
  dplyr::semi_join(x = ., 
                   # join this to the f1units
                   y = f1units %>% 
                     # but filter this to the mean greater than 100
                     dplyr::filter(mean > 100),
                   by = "area")
permits_big %>% dplyr::glimpse(78)
```

Now redo the plot with `ggplot2::geom_line()` with only big cities. 

```{r permits_big_line_grouped}
permits_big %>% 
  ggplot2::ggplot(aes(x = date, y = f1units)) + 
    ggplot2::geom_line(aes(group = area))
```

Add the `alpha = 1/10` to see the lines a little clearer...

```{r permits_big_line_grouped_alpha}
permits_big %>% 
  ggplot2::ggplot(aes(x = date, y = f1units)) + 
    ggplot2::geom_line(aes(group = area), alpha = 1/10)
```

Add the `ggplot2::scale_y_log10()` to reduce the difference in the biggest and smallest `big` cities. 

```{r permits_big_line_grouped_scale_y_log10}
permits_big %>% 
  ggplot2::ggplot(aes(x = date, y = f1units)) + 
    ggplot2::geom_line(aes(group = area), alpha = 1/10) + 
      ggplot2::scale_y_log10()
```

Now we can see a bit of a pattern in the data. 

Add a `ggplot2::geom_smooth()` line to see the long term trends.

```{r geom_smooth}
permits_big %>% 
  ggplot2::ggplot(aes(x = date, y = f1units)) + 
    ggplot2::geom_line(aes(group = area), alpha = 1/10) + 
      ggplot2::scale_y_log10() + 
        ggplot2::geom_smooth(se = FALSE)
```

## Model

Models are a great way to partition the signal into a monthly component. 

Start with one city `"Houston"`.

```{r houston}
houston <- permits %>% dplyr::filter(stringr::str_detect(area, "Houston"))
houston %>% dplyr::count(area)
```

Now plot only Houston.

```{r houston_lineplot}
houston %>% 
  ggplot2::ggplot(aes(x = date, y = f1units)) + 
    ggplot2::geom_line(aes(group = area))
```

Check this again by putting the `month` on the `x` and `group` = `year`. 

```{r group_by_year}
houston %>% 
  ggplot2::ggplot(aes(x = month, y = f1units)) + 
    ggplot2::geom_line(aes(group = year)) + 
      ggplot2::scale_y_log10()
```

This shows more building permits earlier in the year than later in the year. 

Questions:

* is this pattern the same everywhere?

* what drives this? is it the weather?

* Houston in July is less pleasant than Houston in December. 

We build a model.

"I don't believe this is a good model, I'm just going to use it to partition this signal into a monthly effect".

```{r linear_model}
library(modelr)
houston_mod <- lm(log(f1units) ~ factor(month), data = houston)
```

## Look at the predictions

Add the predictions to the model with `modelr::add_predictions()`

```{r add_predictions}
houston %>% 
  # adds a column of predictions to the model
  modelr::add_predictions(houston_mod) %>% 
  ggplot2::ggplot(aes(x = date, y = pred)) + 
  ggplot2::geom_line()
```

The model has captured the seasonal pattern of the permits. 

## Look at the residuals

Now look at the residuals or "what remains after we remove the monthly signal" with `modelr::add_residuals()`.

```{r add_residuals}
houston %>% 
  # adds a column of predictions to the model
  modelr::add_residuals(houston_mod) %>% 
  ggplot2::ggplot(aes(x = date, y = resid)) + 
  ggplot2::geom_line()
```

Now we can see what the pattern is with the monthly pattern removed. 

Questions: 

* What is driving this trend?   
* What happened in 2010?  

## Extend the model to every city

Now use `dplyr`, `tidyr`, and `purrr` to extend the model to every city in the data set. 

The `tidyr::nest()` function works like so:

> `nest()` creates a list of data frames containing all the nested variables: this seems to be the most useful form in practice.

```{r by_area}
by_area <- permits_big %>% 
  dplyr::group_by(area) %>% 
  tidyr::nest()
by_area %>% utils::head(10)
```

This dataset has two columns: `area` is a character vector, and `data` is a column that contains a tibble in each row.

```{r first_tibble}
by_area %>% 
  dplyr::select(data) %>% 
  utils::head(1) %>% 
  utils::str() 
```

`area_model` is the function we will build to extend the model.

```{r area_model}
area_model <- function(df) {
  lm(log10(f1units + 1) ~ factor(x = month), data = df)
}
```

Now we use `purrr::map()` and `purrr::map2()` to extend the model function to each tibble in the `by_area` data frame. 

```{r detrended}
detrended <- by_area %>% dplyr::mutate(
  model = purrr::map(.x = data, .f = area_model),
  # add_residuals is from modelr
  resids = purrr::map2(.x = data, .y = model, .f = add_residuals)
) %>% tidyr::unnest(resids)
detrended %>% glimpse(78)
```

Now we plot the new model data in `resid` column.

```{r plot_detrended}
detrended %>% 
  ggplot2::ggplot(aes(x = date, y = resid)) + 
    ggplot2::geom_line(aes(group = area), alpha = 1/10) + 
        ggplot2::geom_smooth(se = FALSE)
```

This final plot tells us this pattern affects all cities in the data set. 

## Final thoughts 

1. Great problem solving strategy: You have a big problem to solve, start with one small piece (Houston), solve it, then generalize this to the rest of the problem.

2. The entire analysis was motivated by this plot:

```{r important_plot}
permits_big %>% 
  ggplot2::ggplot(aes(x = date, y = f1units)) + 
    ggplot2::geom_line(aes(group = area))
```

After the manipulations, log transformations, and adding `ggplot2::geom_smooth()`, we can build a model to see how much the seasonal pattern contributes to the overall variation. 


